---
title: "04_Real....Anlysis"
author: "Poramed Tripop"
date: "9/30/2021"
output:
  html_document: default
---

```{r setup}
knitr::opts_chunk$set(eval = TRUE)
```

### Introduction and background

This analysis is based on the ......... (found here: https://artscience.blog/home/divvy-dataviz-case-study). The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: “In what ways do members and casual riders use Divvy bikes differently?”

https://divvy-tripdata.s3.amazonaws.com/index.html

### Install required packages
tidyverse for data import and wrangling
lubridate for date functions

```{r install package}
#install.packages("tidyverse")
#install.packages("lubridate")
```

```{r load package}
library(tidyverse)
library(lubridate)
```

```{r displays working directory}
getwd()
#setwd("C:/Users/Action/Documents/RStudio/GGDA/Project/Cyclistic") #sets your working directory to simplify calls to data
```

### STEP 1: COLLECT DATA

#### Upload Cyclistic datasets (csv files) here

https://divvy-tripdata.s3.amazonaws.com/index.html

Divvy_Trips_2019_Q2 is not csv file. So, open it with Excel and save it as csv file before upload to the project.

```{r upload Cyclistic datasets}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

### STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

#### Compare column names each of the files

#### While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file

```{r Compare column names each of the files}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```

#### Rename columns  to make them consisent with q1_2020 (as this will be the supposed going-forward table design for Cyclistic)

```{r rename columns  to make them consisent with q1_2020}
q2_2019 <- rename(q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type")

q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype)

q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype)
```

#### Inspect the dataframes and look for incongruencies

```{r inspect the dataframes and look for incongruencies}
str(q2_2019)
str(q3_2019)
str(q4_2019)
str(q1_2020)
```

#### Convert ride_id and rideable_type to character so that they can stack correctly

```{r}
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))

q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))

q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
```

#### convert started_at and ended_at in q2_2019 from character to datetime

```{r convert started_at and ended_at in q2_2019 to datetime}
q2_2019 <-  mutate(q2_2019, started_at = dmy_hm(started_at, tz = "UTC")
                   ,ended_at = dmy_hm(ended_at, tz = "UTC"))
```

#### Stack individual quarter's data frames into one big data frame

```{r stack 4 data frames into one big data frame}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

#### Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020

```{r Remove lat, long, birthyear, and gender fields}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, 
            birthyear, gender, 
            "01 - Rental Details Duration In Seconds Uncapped", 
            "05 - Member Details Member Birthday Year", 
            "Member Gender", "tripduration"))
```














a# Remove "bad" data

a# The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative

a# We will create a new version of the dataframe (v2) since data is being removed

a# https://www.datasciencemadesimple.com/delete-or-drop-rows-in-r-with-conditions-2/
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]


www END wwwwwwwwwwwwwwwwwwwwwwwwwwwwwwwwww




### STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

#### Inspect the new table that has been created

```{r inspect the new table that has been created}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
tail(all_trips)
str(all_trips)
summary(all_trips)
```

#### There are a few problems we will need to fix:

##### (1) In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.

In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"

Before 2020, Divvy used different labels for these two types of riders ... we will want to make our dataframe consistent with their current nomenclature

N.B.: "Level" is a special property of a column that is retained even if a subset does not contain any values from a specific level

###### Begin by seeing how many observations fall under each usertype

```{r seeing how many observations fall under each usertype}
table(all_trips$member_casual)
```

###### Reassign to the desired values (we will go with the current 2020 labels)

```{r replace "Subscriber" with "member" and "Customer" with "casual"}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))
```

###### Check to make sure the proper number of observations were reassigned
table(all_trips$member_casual)

```{r Check to make sure the proper number of observations were reassigned}
table(all_trips$member_casual)
```

##### (2) The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.

###### Add columns that list the date, year, month, day and day_of_week of each ride

This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level

https://www.statmethods.net/input/dates.html more on date formats in R found at that link

```{r add columns that list the date, year, month, day and day_of_week of each ride}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```



##### (3) We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.

###### Add a "ride_length" calculation to all_trips (in seconds)

https://stat.ethz.ch/R-manual/R-devel/library/base/html/difftime.html

```{r add a "ride_length" calculation to all_trips (in seconds)}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

###### Inspect the structure of the columns

```{r inspedt the structure of the columns}
str(all_trips)
```

###### Convert "ride_length" from Factor to numeric so we can run calculations on the data

```{r Convert "ride_length" from Factor to numeric so we can run calculations on the data}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

##### (4) There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.

###### Remove "bad" data

The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative

https://www.datasciencemadesimple.com/delete-or-drop-rows-in-r-with-conditions-2/

###### We will create a new version of the dataframe (v2) since data is being removed

```{r create a new version of the dataframe (v2) and remove entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | 
                              all_trips$ride_length < 0),]
```
  
### STEP 4: CONDUCT DESCRIPTIVE ANALYSIS

#### Descriptive analysis on ride_length (all figures in seconds)

```{r Descriptive analysis on ride_length (all figures in seconds)}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
```
#### You can condense the four lines above to one line using summary() on the specific attribute

```{r condense the four lines above to one line using summary()}
summary(all_trips_v2$ride_length)
```

#### Compare members and casual users

```{r Compare members and casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

#### See the average ride time by each day for members vs casual users

```{r the average ride time by each day for members, casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + 
            all_trips_v2$day_of_week, FUN = mean)
```

#### Notice that the days of the week are out of order. Let's fix that.

```{r days of the week are out of order,fix that}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, 
                                    levels = c("Sunday", "Monday", 
                                             "Tuesday", "Wednesday", 
                                             "Thursday", "Friday", 
                                             "Saturday"))
```

#### Now, let's run the average ride time by each day for members vs casual users

```{r the average ride time by each day for members vs casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + 
            all_trips_v2$day_of_week, FUN = mean)
```

#### analyze ridership data by type and weekday

```{r analyze ridership data by type and weekday}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n(),							#calculates the number of rides and average duration 
  average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```

#### Let's visualize the number of rides by rider type
  
```{r visualize the number of rides by rider type}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), 
            average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, 
             fill = member_casual)) +
  geom_col(position = "dodge")
```

#### Let's create a visualization for average duration

```{r create a visualization for average duration}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), 
            average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = average_duration, 
             fill = member_casual)) +
  geom_col(position = "dodge")
```

### STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS

#### Create a csv file that we will visualize in Excel, Tableau, or my presentation software

N.B.: This file location is for a Mac. If you are working on a PC, change the file location accordingly (most likely "C:\Users\YOUR_USERNAME\Desktop\...") to export the data. You can read more here:

https://datatofish.com/export-dataframe-to-csv-in-r/

```{r Create a csv file that we will visualize in Excel, Tableau}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + 
                      all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = 'C:/Users/Action/Documents/RStudio/GGDA/Project/Cyclistic/avg_ride_length.csv')    
```

### You're done! Congratulations!


